---
title: "rFIA_PIAL-Demo"
output: html_notebook
---

This is an an [R Markdown](http://rmarkdown.rstudio.com) Notebook documenting the usage of the [rFIA](https://doserlab.com/files/rfia/) package being used to extract density estimates for a certain species of tree, *Pinus albicaulis* (PIAL).

Begin by loading the rFIA package and subsetting to strictly the states of interest (here, states with known native distributions of PIAL). Note that these steps wiil only work if the necessary rFIA data has been downloaded prior to this point.

```{r}
# Load library and navgiate to directory with rFIA data stored in it.
library(rFIA)
fiaDir <- '/RAID1/FIA_Data/allData' ; setwd(fiaDir)
# Subset entire rFIA database to strictly the states of interest (here, states with known PIAL populations)
PIAL <- readFIA(dir=fiaDir, states=c('CA','OR','WA','NV','ID','MT','WY'))
# Clip data to only most recent inventory year
PIAL_MR <- clipFIA(PIAL, mostRecent = TRUE)
```

Now, for each state, generate a density estimate (trees per acre, or TPA) for that state using the most recent (MR) data. We then also convert that estimate to trees per kilometer squared (TPKM values).

```{r stateDemos, echo=TRUE}
# California density
PIAL_CA <- readFIA(dir=fiaDir, states=c('CA'))
# Clip data to only most recent inventory year, then get density estimate
PIAL_CA <- clipFIA(PIAL_CA, mostRecent = TRUE)
PIAL_CA_TPA <- tpa(PIAL_CA, totals = TRUE, bySpecies = TRUE)
CA_TPA <- as.numeric(PIAL_CA_TPA[which(PIAL_CA_TPA$COMMON_NAME=='whitebark pine'),5])
# Calculate density in trees per km2, and report both trees per acre and trees per km2
CA_TPKM <- CA_TPA/0.00404686 ; paste0('CA TPA: ', CA_TPA, '; CA TPKM: ', CA_TPKM)

# Oregon density
PIAL_OR <- readFIA(dir=fiaDir, states=c('OR'))
# Clip data to only most recent inventory year, then get density estimate
PIAL_OR <- clipFIA(PIAL_OR, mostRecent = TRUE)
PIAL_OR_TPA <- tpa(PIAL_OR, totals = TRUE, bySpecies = TRUE)
OR_TPA <- as.numeric(PIAL_OR_TPA[which(PIAL_OR_TPA$COMMON_NAME=='whitebark pine'),5])
# Calculate density in trees per km2, and report both trees per acre and trees per km2
OR_TPKM <- OR_TPA/0.00404686 ; paste0('OR TPA: ', OR_TPA, '; OR TPKM: ', OR_TPKM)

# Washington density
PIAL_WA <- readFIA(dir=fiaDir, states=c('WA'))
# Clip data to only most recent inventory year, then get density estimate
PIAL_WA <- clipFIA(PIAL_WA, mostRecent = TRUE)
PIAL_WA_TPA <- tpa(PIAL_WA, totals = TRUE, bySpecies = TRUE)
WA_TPA <- as.numeric(PIAL_WA_TPA[which(PIAL_WA_TPA$COMMON_NAME=='whitebark pine'),5])
# Calculate density in trees per km2, and report both trees per acre and trees per km2
WA_TPKM <- WA_TPA/0.00404686 ; paste0('WA TPA: ', WA_TPA, '; WA TPKM: ', WA_TPKM)

# Washington density
PIAL_WA <- readFIA(dir=fiaDir, states=c('WA'))
# Clip data to only most recent inventory year, then get density estimate
PIAL_WA <- clipFIA(PIAL_WA, mostRecent = TRUE)
PIAL_WA_TPA <- tpa(PIAL_WA, totals = TRUE, bySpecies = TRUE)
WA_TPA <- as.numeric(PIAL_WA_TPA[which(PIAL_WA_TPA$COMMON_NAME=='whitebark pine'),5])
# Calculate density in trees per km2, and report both trees per acre and trees per km2
WA_TPKM <- WA_TPA/0.00404686 ; paste0('WA TPA: ', WA_TPA, '; WA TPKM: ', WA_TPKM)

# Idaho density
PIAL_ID <- readFIA(dir=fiaDir, states=c('ID'))
# Clip data to only most recent inventory year, then get density estimate
PIAL_ID <- clipFIA(PIAL_ID, mostRecent = TRUE)
PIAL_ID_TPA <- tpa(PIAL_ID, totals = TRUE, bySpecies = TRUE)
ID_TPA <- as.numeric(PIAL_ID_TPA[which(PIAL_ID_TPA$COMMON_NAME=='whitebark pine'),5])
# Calculate density in trees per km2, and report both trees per acre and trees per km2
ID_TPKM <- ID_TPA/0.00404686 ; paste0('ID TPA: ', ID_TPA, '; ID TPKM: ', ID_TPKM)

# Montana density
PIAL_MT <- readFIA(dir=fiaDir, states=c('MT'))
# Clip data to only most recent inventory year, then get density estimate
PIAL_MT <- clipFIA(PIAL_MT, mostRecent = TRUE)
PIAL_MT_TPA <- tpa(PIAL_MT, totals = TRUE, bySpecies = TRUE)
MT_TPA <- as.numeric(PIAL_MT_TPA[which(PIAL_MT_TPA$COMMON_NAME=='whitebark pine'),5])
# Calculate density in trees per km2, and report both trees per acre and trees per km2
MT_TPKM <- MT_TPA/0.00404686 ; paste0('MT TPA: ', MT_TPA, '; MT TPKM: ', MT_TPKM)

# Wyoming density
PIAL_WY <- readFIA(dir=fiaDir, states=c('WY'))
# Clip data to only most recent inventory year, then get density estimate
PIAL_WY <- clipFIA(PIAL_WY, mostRecent = TRUE)
PIAL_WY_TPA <- tpa(PIAL_WY, totals = TRUE, bySpecies = TRUE)
WY_TPA <- as.numeric(PIAL_WY_TPA[which(PIAL_WY_TPA$COMMON_NAME=='whitebark pine'),5])
# Calculate density in trees per km2, and report both trees per acre and trees per km2
WY_TPKM <- WY_TPA/0.00404686 ; paste0('WY TPA: ', WY_TPA, '; WY TPKM: ', WY_TPKM)

```
Finally, one of the GFS populations appear to lie half within TN and half within NC. For that population, calculate the average of the two density estimates from those states.

```{r echo=TRUE}
# Average of WA and ID TPKMs
mean(WA_TPKM, ID_TPKM)
```

```{r echo=TRUE}
# Average of OR and ID TPKMs
mean(OR_TPKM, ID_TPKM)
```

```{r echo=TRUE}
# Average of MT and WY TPKMs
mean(MT_TPKM, WY_TPKM)
```