---
title: "GBIF Occurrence Download & CoordinateCleaner QC Pipeline"
author: "Austin Koontz"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = TRUE, warning = TRUE)
```

# Overview

This document walks through the pipeline used to download and clean plant
species occurrence records from the
[Global Biodiversity Information Facility (GBIF)](https://www.gbif.org/).
The pipeline has two major phases:

1. **Download** — submit a single server-side request to GBIF covering all
   target species, restricted to North America (U.S., Canada, Mexico, and
   the Caribbean), the years 1975–2025, georeferenced records only, and
   excluding fossils and ex situ (botanical garden / zoo) specimens.
2. **Clean** — apply the
   [CoordinateCleaner](https://ropensci.github.io/CoordinateCleaner/)
   package's suite of record-level and dataset-level quality tests to flag
   and remove problematic occurrences, along with a geoprivacy filter
   that removes records whose coordinates were intentionally obscured by
   the data provider.

The pipeline is designed to run unattended in the background on a Linux
server. Expensive stages are checkpointed to `.rds` files so that a
re-run skips any stage whose checkpoint already exists, and certain
steps are parallelized.

## GBIF credentials

`occ_download()` requires a free
[GBIF account](https://www.gbif.org/user/profile). Store your credentials
once in your `.Renviron` so they are picked up automatically:

```r
usethis::edit_r_environ()
# Add these three lines, save, and restart R:
# GBIF_USER=your_username
# GBIF_PWD=your_password
# GBIF_EMAIL=your_email@example.com
```

# Setup

## Load packages

```{r packages}
pacman::p_load(rgbif, taxize, CoordinateCleaner, dplyr, countrycode, 
               readr, purrr, tibble, future, furrr)
```

## Parameters

Key parameters are set here. `coord_uncertainty_km` sets the maximum acceptable
coordinate uncertainty; [CoordinateCleaner](https://ropensci.github.io/CoordinateCleaner/articles/Cleaning_GBIF_data_with_CoordinateCleaner.html) uses 100 km, but for this project 10 km is used.

```{r parameters}
setwd('/home/akoontz/Documents/Indicators/Walder_Indicators/Code/gbifOccurrences/')
species_file <- "speciesLists/SpeciesList_2026-08-20_40n_MexicanSpecies.csv"   # one species name per row, no header
out_dir      <- "gbif_clean_output"
dir.create(out_dir, showWarnings = FALSE)
year_range <- c(1975, 2025)

# North America: US, Canada, Mexico + Caribbean (ISO 3166-1 alpha-2 codes)
countries <- c(
  "US", "CA", "MX",                                            # US, Canada, Mexico
  "AG", "BS", "BB", "CU", "DM", "DO", "GD", "HT", "JM", "KN",   # Caribbean nations
  "LC", "VC", "TT", "AI", "AW", "BQ", "VG", "KY", "CW", "GP",   # + territories
  "MQ", "MS", "PR", "SX", "TC", "VI"
)
# Max acceptable coordinate uncertainty, in km. For range-scale/species-level work,
# a much tighter value like 1-10 km is usually more appropriate.
coord_uncertainty_km <- 10
```

## Parallel plan and logging

This section sets up the number of cores to used for future parallelized steps.

```{r parallel-and-logging}
# ---- PARALLEL PLAN ---------------------------------------------------------------
# Specify number of workers (low number, to leave space on the server)
N_WORKERS <- min(18, future::availableCores() - 2)
plan(multicore, workers = N_WORKERS)   # fork-based parallelization; Linux only
FORCE_RERUN <- FALSE   # set TRUE to ignore existing checkpoints and redo everything

# ---- Logging helper ---------------------------------------------------------------
log_msg <- function(...) {
  message(format(Sys.time(), "[%Y-%m-%d %H:%M:%S] "), ...)
}
script_start <- Sys.time()
pdf(NULL)   # suppress default Rplots.pdf when running non-interactively
log_msg("=== Pipeline started (", N_WORKERS, " parallel workers) ===")
```

# Read species list

The input is a single-column CSV with one species scientific name per row and no header.

```{r read-species}
species_list <- read_csv(species_file, col_names = "species", show_col_types = FALSE) %>%
  pull(species) %>%
  trimws() %>%
  unique()
log_msg(length(species_list), " species read from ", species_file)
```

# Match names to the GBIF backbone taxonomy

Each input name is matched to the
[GBIF Backbone Taxonomy](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c)
via `name_backbone_checklist()`, which returns a numeric `usageKey` for
each name. Names that cannot be matched (`matchType` of `"NONE"` or
`"HIGHERRANK"`) are written to `unmatched_names.csv` for manual review.

GBIF's `taxonKey` occurrence filter automatically includes records filed
under synonyms of a matched accepted name, so no extra synonym-handling
is needed for the download itself.

```{r backbone-match}
log_msg("Matching ", length(species_list), " names to GBIF backbone...")

backbone <- name_backbone_checklist(species_list)

# If there are any unmatched names, write these to a CSV
unmatched <- backbone %>% filter(matchType %in% c("NONE", "HIGHERRANK"))
if (nrow(unmatched) > 0) {
  log_msg(nrow(unmatched), " names did not cleanly match the GBIF backbone; see unmatched_names.csv")
  write_csv(unmatched, file.path(out_dir, "unmatched_names.csv"))
}

accepted <- backbone %>% filter(matchType %in% c("EXACT", "FUZZY"))

log_msg(nrow(accepted), " of ", length(species_list),
        " species matched to a GBIF backbone accepted usage")
```

# Enumerate synonyms (documentation only)

Although GBIF's download filter folds synonyms in automatically, this
step explicitly enumerates each species' known synonyms and writes them
to `synonym_mapping.csv` for reproducibility. Two sources are checked:

1. **GBIF's own backbone** (primary) — via `name_usage(data = "synonyms")`.
2. **Kew's Plants of the World Online** (optional cross-check) — via
   `taxize::synonyms(db = "pow")`, wrapped in `tryCatch` since this
   external API is known to block programmatic access.

```{r enumerate-synonyms}
get_gbif_synonyms <- function(key) {
  tryCatch({
    syn <- name_usage(key = key, data = "synonyms")$data
    if (is.null(syn) || nrow(syn) == 0) return(character(0))
    unique(syn$canonicalName[!is.na(syn$canonicalName)])
  }, error = function(e) character(0))
}

get_pow_synonyms_optional <- function(sp) {
  tryCatch({
    syn_df <- taxize::synonyms(sp, db = "pow")[[1]]
    if (is.null(syn_df) || !is.data.frame(syn_df) || nrow(syn_df) == 0) return(character(0))
    name_col <- intersect(c("name", "fullName", "scientificName"), names(syn_df))[1]
    if (is.na(name_col)) return(character(0))
    unique(as.character(syn_df[[name_col]]))
  }, error = function(e) character(0))
}

log_msg("Enumerating synonyms for ", nrow(accepted), " species (", N_WORKERS, " workers)...")

synonym_table <- accepted %>%
  select(species = verbatim_name, usageKey) %>%
  mutate(
    gbif_synonyms = future_map(usageKey, get_gbif_synonyms, .progress = FALSE),
    pow_synonyms  = future_map(species, get_pow_synonyms_optional, .progress = FALSE)
  )

synonym_report <- synonym_table %>%
  mutate(
    gbif_synonyms = map_chr(gbif_synonyms, ~ paste(.x, collapse = "; ")),
    pow_synonyms  = map_chr(pow_synonyms, ~ paste(.x, collapse = "; "))
  )
write_csv(synonym_report, file.path(out_dir, "synonym_mapping.csv"))

gbif_keys <- unique(accepted$usageKey)
log_msg(length(gbif_keys), " unique GBIF accepted taxon keys will be queried ",
        "(synonym occurrences are included automatically by GBIF's taxonKey filter)")
```

# Download occurrences from GBIF

The download is submitted as a single server-side job via
`occ_download()`. This produces a citable DOI. The download uses the
Darwin Core Archive (`DWCA`) format so the `dataGeneralizations` and 
`informationWithheld` fields are available for the geoprivacy filter 
in the next step.

The following server-side filters are applied:

| Filter | Purpose |
|--------|---------|
| `taxonKey` | Restrict to matched species (synonyms included automatically) |
| `country` | North America + Caribbean |
| `year` | 1975–2025 |
| `hasCoordinate = TRUE` | Only georeferenced records |
| `hasGeospatialIssue = FALSE` | Exclude records GBIF has flagged for geospatial problems |
| `basisOfRecord ≠ FOSSIL_SPECIMEN, LIVING_SPECIMEN` | Exclude fossils and ex situ specimens |

This step is **checkpointed**: if `GBIF_occurrences_raw.rds` already
exists, the download is skipped and cached data is loaded instead.

```{r gbif-download}
raw_rds <- file.path(out_dir, "GBIF_occurrences_raw.rds")

if (file.exists(raw_rds) && !FORCE_RERUN) {
  
  log_msg("Found existing raw download at ", raw_rds, " -- loading from disk (skipping GBIF download)")
  dat_raw <- readRDS(raw_rds)
  
} else {
  
  download_key <- occ_download(
    pred_in("taxonKey", gbif_keys),
    pred_in("country", countries),
    pred_gte("year", year_range[1]),
    pred_lte("year", year_range[2]),
    pred("hasCoordinate", TRUE),
    pred("hasGeospatialIssue", FALSE),
    pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),  # exclude fossils and ex situ (zoo/garden) specimens
    format = "DWCA"
  )
  log_msg("Download submitted: ", download_key,
          " -- waiting for GBIF to prepare it (status printed every 60s)...")
  
  # status_ping controls how often occ_download_wait() polls and prints status,
  # so background progress is visible in run_log.txt
  occ_download_wait(download_key, status_ping = 60, quiet = FALSE)
  dl_path <- occ_download_get(download_key, path = out_dir, overwrite = TRUE)
  dat_raw <- occ_download_import(dl_path)
  log_msg(nrow(dat_raw), " raw occurrence records downloaded")
  saveRDS(dat_raw, raw_rds)
  # Save the citation 
  writeLines(format(gbif_citation(dl_path)), file.path(out_dir, "gbif_citation.txt"))
}
```

## Identify synonym renames

GBIF labels every downloaded record with its **accepted** species name,
not the queried name. If an input name is a synonym in GBIF's backbone,
its records appear under a different name in the `species` column. This
step detects those cases and traces each "extra" accepted name back to
the input synonym via shared `usageKey` values. The resulting
`name_lookup` table is used later to label output filenames with both
names.

```{r synonym-lookup}
# Build lookup: input name -> GBIF accepted name, by comparing the input
# species list against the species names that actually appear in the downloaded
# data. GBIF's download labels records with its accepted species name, so any
# input name that doesn't appear verbatim in dat_raw$species was a synonym.
gbif_species <- unique(dat_raw$species)
extra_species <- setdiff(gbif_species, species_list)   # names in download but not in input
missing_input <- setdiff(species_list, gbif_species)   # input names not in download (synonym or no records)

# Match each extra (accepted) name back to the input (synonym) name via the
# backbone's usageKey: both the input synonym and the accepted name share the
# same accepted usageKey
name_lookup <- tibble(input_name = character(), gbif_name = character())
if (length(extra_species) > 0 && length(missing_input) > 0) {
  # For each extra species, look up its usageKey, then find which input name
  # mapped to the same key
  extra_keys <- map_dfr(extra_species, function(sp) {
    bb <- name_backbone(name = sp, kingdom = "Plantae")
    tibble(gbif_name = sp, key = bb$usageKey)
  })
  input_keys <- accepted %>%
    transmute(input_name = verbatim_name, key = usageKey)
  name_lookup <- inner_join(input_keys, extra_keys, by = "key") %>%
    filter(input_name != gbif_name) %>%
    select(input_name, gbif_name) %>%
    distinct()
}

if (nrow(name_lookup) > 0) {
  log_msg(nrow(name_lookup), " input names are synonyms in GBIF's backbone:")
  for (i in seq_len(nrow(name_lookup))) {
    log_msg("  ", name_lookup$input_name[i], " -> ", name_lookup$gbif_name[i])
  }
}
```

# Prepare data for cleaning

The raw download is trimmed to the columns needed for cleaning and
filtering, and records lacking coordinates are dropped.

```{r prepare-data}
dat <- dat_raw %>%
  select(species, decimalLongitude, decimalLatitude, countryCode,
         individualCount, gbifID, family, taxonRank,
         coordinateUncertaintyInMeters, year, basisOfRecord,
         institutionCode, datasetKey,
         dataGeneralizations, informationWithheld) %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude))
```

## Geoprivacy filter

Some data providers intentionally obscure, randomize, or generalize
coordinates for sensitive species. This step uses pattern matching on 
the `dataGeneralizations` and `informationWithheld` fields to identify 
and remove such records.

A per-species tally of removed records is written to
`geoprivacy_removed_by_species.csv`.

```{r geoprivacy-filter}
# %%% 6b. Filter records with obscured/generalized coordinates for geoprivacy ---
# Matches common phrases data providers use when coordinates have been
# randomized, generalized, or obscured for sensitive-species protection.
geoprivacy_pattern <- paste(
  "coordinate.*obscur", "coordinate.*generaliz", "coordinate.*randomiz",
  "coordinate.*round", "coordinate.*truncat", "coordinate.*alter",
  "location.*obscur", "location.*generaliz", "location.*randomiz",
  "location.*withheld", "geoprivacy", "sensitive.*species",
  sep = "|"
)
dat$geoprivacy_flag <- grepl(geoprivacy_pattern, dat$dataGeneralizations, ignore.case = TRUE) |
  grepl(geoprivacy_pattern, dat$informationWithheld, ignore.case = TRUE)

geoprivacy_tally <- dat %>%
  group_by(species) %>%
  summarise(
    total_records     = n(),
    geoprivacy_removed = sum(geoprivacy_flag),
    .groups = "drop"
  ) %>%
  filter(geoprivacy_removed > 0)

if (nrow(geoprivacy_tally) > 0) {
  log_msg(sum(dat$geoprivacy_flag), " records flagged for geoprivacy coordinate obscuring across ",
          nrow(geoprivacy_tally), " species")
  write_csv(geoprivacy_tally, file.path(out_dir, "geoprivacy_removed_by_species.csv"))
} else {
  log_msg("No records flagged for geoprivacy coordinate obscuring")
}
dat <- dat %>%
  filter(!geoprivacy_flag) %>%
  select(-dataGeneralizations, -informationWithheld, -geoprivacy_flag)
```

## Country code conversion and indexing

Country codes are converted from ISO 3166-1 alpha-2 (as GBIF provides)
to alpha-3 (as `cc_coun()` requires), and a stable `row_id` is added
for realigning parallel chunk results later.

```{r country-codes}
# ISO2 -> ISO3, required by cc_coun()
dat$countryCode <- countrycode(dat$countryCode, origin = "iso2c", destination = "iso3c")

dat <- as.data.frame(dat)
dat$row_id <- seq_len(nrow(dat))   # stable index, used to realign parallel chunk results

log_msg(nrow(dat), " records ready for cleaning")
```

# CoordinateCleaner: record-level tests

The [CoordinateCleaner](https://ropensci.github.io/CoordinateCleaner/)
package provides automated tests for common problems in species
occurrence coordinates. 

| Function | What it flags |
|----------|---------------|
| `cc_val()` | Invalid coordinates (outside valid lat/lon range) |
| `cc_equ()` | Identical latitude and longitude (often a data-entry error) |
| `cc_cap()` | Coordinates matching country or province capitals |
| `cc_cen()` | Coordinates matching country or province centroids |
| `cc_coun()` | Coordinates that fall outside the stated country |
| `cc_sea()` | Coordinates that fall in the ocean |
| `cc_zero()` | Coordinates at (0, 0) — the "null island" default |
| `cc_dupl()` | Exact duplicate records (same species, same coordinates) |

```{r cc-record-level}
log_msg("Running record-level CoordinateCleaner tests...")

clean <- dat %>%
  cc_val() %>%
  cc_equ() %>%
  cc_cap() %>%
  cc_cen() %>%
  cc_coun(iso3 = "countryCode") %>%
  cc_sea() %>%
  cc_zero() %>%
  cc_dupl()

log_msg(nrow(dat) - nrow(clean), " records removed by record-level tests; ", nrow(clean), " remaining")
```

## Geographic outlier detection (`cc_outl`)

`cc_outl()` flags records that are statistical geographic outliers
relative to other records for the same species. It operates per species
(computing pairwise distances). This step is **checkpointed** to `cc_outl_flags.rds`.

```{r cc-outl}
outl_rds <- file.path(out_dir, "cc_outl_flags.rds")

if (file.exists(outl_rds) && !FORCE_RERUN) {
  log_msg("Found existing cc_outl() flags at ", outl_rds, " -- loading from disk")
  outl_flags <- readRDS(outl_rds)
} else {
  log_msg("Running cc_outl() across ", length(unique(clean$species)),
          " species (", N_WORKERS, " workers)...")

  species_chunks <- split(unique(clean$species),
                          cut(seq_along(unique(clean$species)), N_WORKERS, labels = FALSE))
  
  outl_results <- future_map(species_chunks, function(sp_chunk) {
    sub <- clean[clean$species %in% sp_chunk, ]
    flags <- cc_outl(sub, species = "species", value = "flagged", verbose = FALSE)
    tibble(row_id = sub$row_id, outl_flag = flags)
  }, .progress = FALSE)
  outl_flags <- bind_rows(outl_results) %>% arrange(row_id)
  saveRDS(outl_flags, outl_rds)
}
clean <- clean %>%
  left_join(outl_flags, by = "row_id") %>%
  filter(outl_flag) %>%
  select(-outl_flag)
log_msg(nrow(clean), " records remaining after cc_outl()")
```

# Coordinate precision filter

This step removes any record whose
reported `coordinateUncertaintyInMeters` exceeds the threshold set in
the parameters above (`coord_uncertainty_km` km). Records with no
reported uncertainty (`NA`) are retained, following the convention in the
CoordinateCleaner tutorial; these should be inspected separately later on.

```{r precision-filter}
clean <- clean %>%
  filter(is.na(coordinateUncertaintyInMeters) |
           (coordinateUncertaintyInMeters / 1000) <= coord_uncertainty_km)
log_msg(nrow(clean), " records remaining after precision filter (<= ",
        coord_uncertainty_km, " km)")
```

# CoordinateCleaner: dataset-level tests

These tests evaluate coordinate patterns across entire GBIF datasets
(identified by `datasetKey`), flagging whole datasets rather than
individual records.

**`cd_ddmm()`** tests whether a dataset's coordinate distribution is
consistent with a systematic "dd.mm → dd.dd" conversion error — a
common data-entry mistake where degree-minute values are incorrectly
recorded as decimal degrees.

**`cd_round()`** tests whether a dataset's coordinates show an unusually
regular lattice pattern, indicating that points were assigned to
grid-cell centroids rather than actual sampling locations.

Diagnostic reports are written to
`ddmm_conversion_flags_by_dataset.csv` and
`rasterized_sampling_flags_by_dataset.csv` for manual review.
This step is **checkpointed** to `dataset_level_flags.rds`.

```{r dataset-level-tests}
dataset_rds <- file.path(out_dir, "dataset_level_flags.rds")
if (file.exists(dataset_rds) && !FORCE_RERUN) {
  log_msg("Found existing dataset-level flags at ", dataset_rds, " -- loading from disk")
  saved <- readRDS(dataset_rds)
  ddmm_report  <- saved$ddmm_report
  round_report <- saved$round_report
  ddmm_flags   <- saved$ddmm_flags
  round_flags  <- saved$round_flags
} else {
  log_msg("Running cd_ddmm() and cd_round() across ", length(unique(clean$datasetKey)),
          " datasets (", N_WORKERS, " workers)...")
  dataset_ids <- unique(clean$datasetKey)
  dataset_chunks <- split(dataset_ids,
                          cut(seq_along(dataset_ids), N_WORKERS, labels = FALSE))
  
  chunk_results <- future_map(dataset_chunks, function(ds_chunk) {
    sub <- clean[clean$datasetKey %in% ds_chunk, ]
    
    # tryCatch: cd_ddmm() can fail on datasets with very few records (binom.test
    # receives invalid n < x). Treat such chunks as passing (all TRUE).
    ddmm_ds <- tryCatch({
      res <- cd_ddmm(sub, lon = "decimalLongitude", lat = "decimalLatitude",
                     ds = "datasetKey", diagnostic = TRUE, diff = 1, value = "dataset")
      tibble::rownames_to_column(res, var = "datasetKey")
    }, error = function(e) {
      message("  cd_ddmm() failed for a chunk (", length(ds_chunk), " datasets): ", conditionMessage(e))
      NULL
    })
    ddmm_fl <- tryCatch({
      cd_ddmm(sub, lon = "decimalLongitude", lat = "decimalLatitude",
              ds = "datasetKey", diff = 1, value = "flagged")
    }, error = function(e) rep(TRUE, nrow(sub)))
    
    round_ds <- tryCatch({
      cd_round(sub, lon = "decimalLongitude", lat = "decimalLatitude",
               ds = "datasetKey", T1 = 7, value = "dataset", graphs = FALSE)
    }, error = function(e) {
      message("  cd_round() failed for a chunk (", length(ds_chunk), " datasets): ", conditionMessage(e))
      NULL
    })
    round_fl <- tryCatch({
      cd_round(sub, lon = "decimalLongitude", lat = "decimalLatitude",
               ds = "datasetKey", T1 = 7, value = "flagged", graphs = FALSE)
    }, error = function(e) rep(TRUE, nrow(sub)))
    
    list(
      ddmm_report  = ddmm_ds,
      round_report = round_ds,
      flags        = tibble(row_id = sub$row_id, ddmm_flag = ddmm_fl, round_flag = round_fl)
    )
  }, .progress = FALSE)
  
  ddmm_report  <- bind_rows(map(chunk_results, "ddmm_report"))
  round_report <- bind_rows(map(chunk_results, "round_report"))
  flags_all    <- bind_rows(map(chunk_results, "flags")) %>% arrange(row_id)
  ddmm_flags   <- flags_all$ddmm_flag
  round_flags  <- flags_all$round_flag
  
  write_csv(ddmm_report, file.path(out_dir, "ddmm_conversion_flags_by_dataset.csv"))
  write_csv(round_report, file.path(out_dir, "rasterized_sampling_flags_by_dataset.csv"))
  
  saveRDS(list(ddmm_report = ddmm_report, round_report = round_report,
               ddmm_flags = ddmm_flags, round_flags = round_flags),
          dataset_rds)
}

# Remove records belonging to datasets flagged by either test.
# NOTE: these are dataset-level tests, so review ddmm_conversion_flags_by_dataset.csv
# and rasterized_sampling_flags_by_dataset.csv before treating this as final -
# a flagged dataset may still contain a mix of good and bad records.
clean_final <- clean[ddmm_flags & round_flags, ]
log_msg(nrow(clean) - nrow(clean_final),
        " records removed for belonging to a flagged dataset (ddmm/rasterized tests)")
log_msg(nrow(clean_final), " final cleaned records")
```

# Save outputs

Cleaned records are written to one CSV per species in
`gbif_clean_output/species_csvs/`. Coordinate column names are
lowercased to `decimallongitude` and `decimallatitude` for compatibility
with Genes from Space. For synonym cases where GBIF's accepted name
differs from the input name, the output filename includes both:
`InputName_AcceptedName_Nn_YYYY-MM-DD.csv`.

```{r save-outputs}
# Rename column names, into a format more useful for GFS
clean_final <- clean_final %>%
  select(-row_id) %>%
  rename(decimallongitude = decimalLongitude, decimallatitude = decimalLatitude)
species_dir <- file.path(out_dir, "species_csvs")
dir.create(species_dir, showWarnings = FALSE)

for (sp in unique(clean_final$species)) {
  sp_data <- clean_final[clean_final$species == sp, ]
  
  # If GBIF's accepted name differs from the input name, prepend the input name
  input_match <- name_lookup$input_name[name_lookup$gbif_name == sp]
  if (length(input_match) == 1 && input_match != sp) {
    file_prefix <- paste0(gsub(" ", "_", input_match), "_", gsub(" ", "_", sp))
  } else {
    file_prefix <- gsub(" ", "_", sp)
  }
  sp_file <- file.path(species_dir, paste0(file_prefix, "_", nrow(sp_data), "n_",
                                           format(Sys.Date(), "%Y-%m-%d"), ".csv"))
  write_csv(sp_data, sp_file)
}
elapsed <- round(difftime(Sys.time(), script_start, units = "mins"), 1)
log_msg("=== Done in ", elapsed, " minutes. Cleaned data written to: ",
        species_dir, " (one CSV per species) ===")
```

# Reconciliation: missing species

Not every input species will have records in the final output. This
section identifies which species are missing and diagnoses why by
checking whether they were present at each pipeline stage. Species whose
input name was a synonym (with records present under the GBIF accepted
name) are excluded from the "missing" count.

```{r reconciliation}
# Reconcile: which input species have no records in the final output, and why?
# Exclude synonym species that DO have output under their accepted name
synonym_inputs <- name_lookup$input_name[name_lookup$gbif_name %in% unique(clean_final$species)]
missing_spp <- setdiff(species_list, c(unique(clean_final$species), synonym_inputs))
if (length(missing_spp) > 0) {
  log_msg(length(missing_spp), " species from the input list have no final output:")
  missing_info <- tibble(species = missing_spp) %>%
    mutate(
      in_gbif_download  = species %in% unique(dat_raw$species),
      after_cc_records   = species %in% unique(dat$species),
      after_cc_pipeline = species %in% unique(clean$species),
      after_precision   = species %in% unique(clean_final$species)
    )
  for (i in seq_len(nrow(missing_info))) {
    row <- missing_info[i, ]
    if (!row$in_gbif_download) {
      reason <- "no records returned by GBIF download (may be a synonym; check synonym_mapping.csv)"
    } else if (!row$after_cc_records) {
      reason <- "all records lacked coordinates"
    } else if (!row$after_cc_pipeline) {
      reason <- "all records removed by CoordinateCleaner tests (cc_val/cc_equ/.../cc_outl)"
    } else {
      reason <- "all records removed by precision or dataset-level filters"
    }
    log_msg("  - ", row$species, ": ", reason)
  }
  write_csv(missing_info, file.path(out_dir, "species_missing_from_output.csv"))
} else {
  log_msg("All ", length(species_list), " input species have records in the final output")
}
```
