---
title: "GBIF Occurrence Download & CoordinateCleaner QC Pipeline"
author: "Austin Koontz"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: flatly
    code_folding: show
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = FALSE, message = TRUE, warning = TRUE)
```

# Overview

This document walks through the pipeline used to download and clean occurrence records from the
[Global Biodiversity Information Facility (GBIF)](https://www.gbif.org/).
The pipeline has two major phases:

1. **Download** — submit a single server-side request to GBIF covering all
   target species, restricted to specified countries (currently, North America:
   U.S., Canada, Mexico, and the Caribbean), the specified years (currently, 1975–2025), 
   georeferenced records only, and excluding fossils and ex situ (botanical garden / zoo) specimens.
2. **Clean** — apply the
   [CoordinateCleaner](https://ropensci.github.io/CoordinateCleaner/)
   package's suite of record-level and dataset-level quality tests to flag
   and remove problematic occurrences.

The pipeline is designed to run unattended in the background (on a Linux
server). Expensive stages are checkpointed to `.rds` files so that a
re-run skips any stage whose checkpoint already exists. Parallelization
(via the `future` / `furrr` packages) is used for the three most
computationally expensive steps: geographic outlier detection (`cc_outl`),
degree-minute conversion error detection (`cd_ddmm`), and rasterized
sampling detection (`cd_round`).

**Note**: even after running this script, it's worth checking the resulting occurrences to make sure they've
been filtered correctly. From testing, ex situ and other aberrant occurrences can still remain for a species
even after it's been filtered using this script.

## GBIF credentials

`occ_download()` requires a free
[GBIF account](https://www.gbif.org/user/profile). Store your credentials
once in your `.Renviron` so they are picked up automatically:

```r
usethis::edit_r_environ()
# Add these three lines, save, and restart R:
# GBIF_USER=your_username
# GBIF_PWD=your_password
# GBIF_EMAIL=your_email@example.com
```

# Setup

## Load packages

```{r packages}
pacman::p_load(rgbif, taxize, CoordinateCleaner, dplyr, countrycode,
               readr, purrr, tibble, future, furrr)
```

## Parameters

Key parameters that control the scope of the download and the strictness
of cleaning are set here so they can be adjusted in one place.

- **`species_file`**: a single-column CSV of species scientific names (no header).
- **`year_range`**: only occurrences recorded between these years are
  downloaded.
- **`countries`**: ISO 3166-1 alpha-2 codes defining the geographic scope.
  The list below covers the U.S., Canada, Mexico, and Caribbean nations
  and territories.
- **`coord_uncertainty_km`**: maximum acceptable coordinate uncertainty.
  Records reporting uncertainty above this threshold are removed. The
  [CoordinateCleaner tutorial](https://ropensci.github.io/CoordinateCleaner/articles/Cleaning_GBIF_data_with_CoordinateCleaner.html)
  uses 100 km for a global macroecological grain; for range-scale or
  species-level conservation genetics work, 10 km is a more appropriate
  starting point.

```{r parameters}
setwd('/home/akoontz/Documents/Indicators/Walder_Indicators/Scripts/GBIF_occurrences/')
species_file <- "speciesLists/SpeciesList_2026-08-20_40n_MexicanSpecies.csv"
out_dir      <- "gbif_clean_output"
dir.create(out_dir, showWarnings = FALSE)

year_range <- c(1975, 2025)

countries <- c(
  "US", "CA", "MX",
  "AG", "BS", "BB", "CU", "DM", "DO", "GD", "HT", "JM", "KN",
  "LC", "VC", "TT", "AI", "AW", "BQ", "VG", "KY", "CW", "GP",
  "MQ", "MS", "PR", "SX", "TC", "VI"
)

coord_uncertainty_km <- 10
```

## Parallel plan and logging

The parallel plan is conservative by design: it uses fewer workers than
available cores to leave headroom on a shared server and avoid
memory-exhaustion crashes on spatial workloads. `plan(multicore)` uses
fork-based parallelism (Linux only); on Windows or inside RStudio, use
`plan(multisession)` instead.

```{r parallel-and-logging}
N_WORKERS <- min(18, future::availableCores() - 2)
plan(multicore, workers = N_WORKERS)
FORCE_RERUN <- FALSE

log_msg <- function(...) {
  message(format(Sys.time(), "[%Y-%m-%d %H:%M:%S] "), ...)
}
script_start <- Sys.time()
pdf(NULL)
log_msg("=== Pipeline started (", N_WORKERS, " parallel workers) ===")
```

# Read species list

The input is a simple CSV with one species name per row and no header.

```{r read-species}
species_list <- read_csv(species_file, col_names = "species", show_col_types = FALSE) %>%
  pull(species) %>%
  trimws() %>%
  unique()
log_msg(length(species_list), " species read from ", species_file)
```

# Match names to the GBIF backbone taxonomy {#backbone}

Each input name is matched to the
[GBIF Backbone Taxonomy](https://www.gbif.org/dataset/d7dddbf4-2cf0-4f39-9b2a-bb099caae36c)
using `name_backbone_checklist()`. This returns a `usageKey` — a unique
numeric identifier — for each name. Names that cannot be matched
(`matchType` of `"NONE"` or `"HIGHERRANK"`) are written to
`unmatched_names.csv` for manual review.

An important detail: GBIF's `taxonKey` occurrence filter automatically
includes records filed under synonyms of a matched accepted name. So once
a species is matched to its accepted backbone key, the download in the
next step already captures records that were originally filed under
older or alternate names — no extra synonym-handling step is needed for
the download itself.

```{r backbone-match}
log_msg("Matching ", length(species_list), " names to GBIF backbone...")

backbone <- name_backbone_checklist(species_list)

unmatched <- backbone %>% filter(matchType %in% c("NONE", "HIGHERRANK"))
if (nrow(unmatched) > 0) {
  log_msg(nrow(unmatched), " names did not cleanly match the GBIF backbone; see unmatched_names.csv")
  write_csv(unmatched, file.path(out_dir, "unmatched_names.csv"))
}

accepted <- backbone %>% filter(matchType %in% c("EXACT", "FUZZY"))

log_msg(nrow(accepted), " of ", length(species_list),
        " species matched to a GBIF backbone accepted usage")
```

# Enumerate synonyms (documentation only)

Although GBIF's download filter already folds synonyms in automatically,
this step explicitly enumerates each species' known synonyms and writes
them to `synonym_mapping.csv`. This serves as documentation: it provides
a record of exactly which alternate names were folded into each query,
which is useful for reproducibility and for understanding why a species'
occurrence count might be higher than expected.

Two sources are checked:

1. **GBIF's own backbone** (primary) — via `name_usage(data = "synonyms")`.
   This is the same taxonomy that drives the download, so it is authoritative.
2. **Kew's Plants of the World Online** (optional cross-check) — via
   `taxize::synonyms(db = "pow")`. This external API is known to
   rate-limit or block programmatic access (HTTP 403/429/502), so the
   call is wrapped in `tryCatch` and failures are silently skipped.

```{r enumerate-synonyms}
get_gbif_synonyms <- function(key) {
  tryCatch({
    syn <- name_usage(key = key, data = "synonyms")$data
    if (is.null(syn) || nrow(syn) == 0) return(character(0))
    unique(syn$canonicalName[!is.na(syn$canonicalName)])
  }, error = function(e) character(0))
}

get_pow_synonyms_optional <- function(sp) {
  tryCatch({
    syn_df <- taxize::synonyms(sp, db = "pow")[[1]]
    if (is.null(syn_df) || !is.data.frame(syn_df) || nrow(syn_df) == 0) return(character(0))
    name_col <- intersect(c("name", "fullName", "scientificName"), names(syn_df))[1]
    if (is.na(name_col)) return(character(0))
    unique(as.character(syn_df[[name_col]]))
  }, error = function(e) character(0))
}

log_msg("Enumerating synonyms for ", nrow(accepted), " species (", N_WORKERS, " workers)...")

synonym_table <- accepted %>%
  select(species = verbatim_name, usageKey) %>%
  mutate(
    gbif_synonyms = future_map(usageKey, get_gbif_synonyms, .progress = FALSE),
    pow_synonyms  = future_map(species, get_pow_synonyms_optional, .progress = FALSE)
  )

synonym_report <- synonym_table %>%
  mutate(
    gbif_synonyms = map_chr(gbif_synonyms, ~ paste(.x, collapse = "; ")),
    pow_synonyms  = map_chr(pow_synonyms, ~ paste(.x, collapse = "; "))
  )
write_csv(synonym_report, file.path(out_dir, "synonym_mapping.csv"))

gbif_keys <- unique(accepted$usageKey)
log_msg(length(gbif_keys), " unique GBIF accepted taxon keys will be queried ",
        "(synonym occurrences are included automatically by GBIF's taxonKey filter)")
```

# Download occurrences from GBIF

The download is submitted as a single server-side job via `occ_download()`.
This is preferable to `occ_search()` or `occ_data()` for several reasons:

- **No record-count cap**: `occ_search()` is limited to 100,000 records
  per call; `occ_download()` has no such limit.
- **Citable DOI**: each download receives a DOI that should be cited in
  any resulting publication.
- **Reproducibility**: the download is logged on GBIF's servers and can be
  re-accessed later.

The following server-side filters are applied:

| Filter | Purpose |
|--------|---------|
| `taxonKey` | Restrict to the matched species (synonyms included automatically) |
| `country` | Restrict to North America + Caribbean |
| `year` | 1975–2025 |
| `hasCoordinate = TRUE` | Only georeferenced records |
| `hasGeospatialIssue = FALSE` | Exclude records GBIF has flagged for geospatial problems |
| `basisOfRecord ≠ FOSSIL_SPECIMEN, LIVING_SPECIMEN` | Exclude fossils and ex situ (zoo/garden) specimens |

This step is **checkpointed**: if the `GBIF_occurrences_raw.rds` file already
exists, the download is skipped and the cached data is loaded instead.

```{r gbif-download}
raw_rds <- file.path(out_dir, "GBIF_occurrences_raw.rds")

if (file.exists(raw_rds) && !FORCE_RERUN) {

  log_msg("Found existing raw download at ", raw_rds, " -- loading from disk (skipping GBIF download)")
  dat_raw <- readRDS(raw_rds)

} else {

  download_key <- occ_download(
    pred_in("taxonKey", gbif_keys),
    pred_in("country", countries),
    pred_gte("year", year_range[1]),
    pred_lte("year", year_range[2]),
    pred("hasCoordinate", TRUE),
    pred("hasGeospatialIssue", FALSE),
    pred_not(pred_in("basisOfRecord", c("FOSSIL_SPECIMEN", "LIVING_SPECIMEN"))),
    format = "SIMPLE_CSV"
  )

  log_msg("Download submitted: ", download_key,
          " -- waiting for GBIF to prepare it (status printed every 60s)...")

  occ_download_wait(download_key, status_ping = 60, quiet = FALSE)

  dl_path <- occ_download_get(download_key, path = out_dir, overwrite = TRUE)
  dat_raw <- occ_download_import(dl_path)

  log_msg(nrow(dat_raw), " raw occurrence records downloaded")

  saveRDS(dat_raw, raw_rds)

  writeLines(format(gbif_citation(dl_path)), file.path(out_dir, "gbif_citation.txt"))
}
```

## Identify synonym renames

GBIF labels every downloaded record with its **accepted** species name,
not necessarily the name that was queried. If an input name is treated as
a synonym by GBIF's backbone, its records will appear under a different
name in the `species` column. This step detects those cases by comparing
the input species list against the names actually present in the download,
and traces each "extra" accepted name back to the input synonym via
shared `usageKey` values.

The resulting `name_lookup` table is used later to label output CSV
filenames with both the original input name and the GBIF accepted name
(e.g., `Hesperocyparis_forbesii_Cupressus_arizonica_123n_2026-08-20.csv`).

```{r synonym-lookup}
gbif_species <- unique(dat_raw$species)
extra_species <- setdiff(gbif_species, species_list)
missing_input <- setdiff(species_list, gbif_species)

name_lookup <- tibble(input_name = character(), gbif_name = character())
if (length(extra_species) > 0 && length(missing_input) > 0) {
  extra_keys <- map_dfr(extra_species, function(sp) {
    bb <- name_backbone(name = sp, kingdom = "Plantae")
    tibble(gbif_name = sp, key = bb$usageKey)
  })
  input_keys <- accepted %>%
    transmute(input_name = verbatim_name, key = usageKey)
  name_lookup <- inner_join(input_keys, extra_keys, by = "key") %>%
    filter(input_name != gbif_name) %>%
    select(input_name, gbif_name) %>%
    distinct()
}

if (nrow(name_lookup) > 0) {
  log_msg(nrow(name_lookup), " input names are synonyms in GBIF's backbone:")
  for (i in seq_len(nrow(name_lookup))) {
    log_msg("  ", name_lookup$input_name[i], " -> ", name_lookup$gbif_name[i])
  }
}
```

# Prepare data for cleaning

Before running `CoordinateCleaner`, the raw download is trimmed to the
columns needed for cleaning, records lacking coordinates are dropped,
and country codes are converted from ISO 3166-1 alpha-2 (as GBIF provides
them) to alpha-3 (as `cc_coun()` requires).

A stable `row_id` column is added to allow parallel chunk results to be
realigned with the full table after processing.

```{r prepare-data}
dat <- dat_raw %>%
  select(species, decimalLongitude, decimalLatitude, countryCode,
         individualCount, gbifID, family, taxonRank,
         coordinateUncertaintyInMeters, year, basisOfRecord,
         institutionCode, datasetKey) %>%
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude))

dat$countryCode <- countrycode(dat$countryCode, origin = "iso2c", destination = "iso3c")

dat <- as.data.frame(dat)
dat$row_id <- seq_len(nrow(dat))

log_msg(nrow(dat), " records ready for cleaning")
```

# CoordinateCleaner: record-level tests

The [CoordinateCleaner](https://ropensci.github.io/CoordinateCleaner/)
package provides a suite of automated tests for flagging common problems
in species occurrence coordinates. Each `cc_*` function removes records 
that fail its test. The tests, in order:

| Function | What it flags |
|----------|---------------|
| `cc_val()` | Invalid coordinates (outside valid lat/lon range) |
| `cc_equ()` | Identical latitude and longitude (often a data-entry error) |
| `cc_cap()` | Coordinates matching country or province capitals |
| `cc_cen()` | Coordinates matching country or province centroids |
| `cc_coun()` | Coordinates that fall outside the stated country |
| `cc_sea()` | Coordinates that fall in the ocean |
| `cc_zero()` | Coordinates at (0, 0) — the "null island" default |
| `cc_dupl()` | Exact duplicate records (same species, same coordinates) |

```{r cc-record-level}
log_msg("Running record-level CoordinateCleaner tests...")

clean <- dat %>%
  cc_val() %>%
  cc_equ() %>%
  cc_cap() %>%
  cc_cen() %>%
  cc_coun(iso3 = "countryCode") %>%
  cc_sea() %>%
  cc_zero() %>%
  cc_dupl()

log_msg(nrow(dat) - nrow(clean), " records removed by record-level tests; ", nrow(clean), " remaining")
```

## Geographic outlier detection (`cc_outl`)

`cc_outl()` flags records that are statistical geographic outliers
relative to the other records for the same species. Unlike the tests
above, this one operates **per species** (it computes pairwise distances
among each species' points), which makes it both the most expensive
record-level test and a good candidate for parallelization.

The species list is split into chunks, and each chunk is processed by a
separate worker. Because `cc_outl()` treats each species independently,
splitting by species and recombining gives identical results to running
the full table at once.

This step is **checkpointed**: if `cc_outl_flags.rds` already exists, the
saved flags are loaded instead of re-computing.

```{r cc-outl}
outl_rds <- file.path(out_dir, "cc_outl_flags.rds")

if (file.exists(outl_rds) && !FORCE_RERUN) {

  log_msg("Found existing cc_outl() flags at ", outl_rds, " -- loading from disk")
  outl_flags <- readRDS(outl_rds)

} else {

  log_msg("Running cc_outl() across ", length(unique(clean$species)),
          " species (", N_WORKERS, " workers)...")

  species_chunks <- split(unique(clean$species),
                          cut(seq_along(unique(clean$species)), N_WORKERS, labels = FALSE))

  outl_results <- future_map(species_chunks, function(sp_chunk) {
    sub <- clean[clean$species %in% sp_chunk, ]
    flags <- cc_outl(sub, species = "species", value = "flagged", verbose = FALSE)
    tibble(row_id = sub$row_id, outl_flag = flags)
  }, .progress = FALSE)

  outl_flags <- bind_rows(outl_results) %>% arrange(row_id)
  saveRDS(outl_flags, outl_rds)
}

clean <- clean %>%
  left_join(outl_flags, by = "row_id") %>%
  filter(outl_flag) %>%
  select(-outl_flag)

log_msg(nrow(clean), " records remaining after cc_outl()")
```

# Coordinate precision filter

Records with very high coordinate uncertainty are unlikely to be useful
for downstream spatial analyses. This step removes any record whose
reported `coordinateUncertaintyInMeters` exceeds the threshold set in
the parameters above (`coord_uncertainty_km` km).

Records with **no** reported uncertainty (`NA`) are retained, following
the convention in the
[CoordinateCleaner tutorial](https://ropensci.github.io/CoordinateCleaner/articles/Cleaning_GBIF_data_with_CoordinateCleaner.html).
These should be inspected separately if coordinate precision matters for
the downstream analysis.

```{r precision-filter}
clean <- clean %>%
  filter(is.na(coordinateUncertaintyInMeters) |
           (coordinateUncertaintyInMeters / 1000) <= coord_uncertainty_km)

log_msg(nrow(clean), " records remaining after precision filter (<= ",
        coord_uncertainty_km, " km)")
```

# CoordinateCleaner: dataset-level tests

Unlike the record-level tests above, these two tests evaluate patterns
across entire **datasets** (identified by `datasetKey`). They flag whole
datasets rather than individual records, so their diagnostic reports
(written to CSV) should be reviewed manually before accepting the
automatic filtering.

## Degree-minute conversion errors (`cd_ddmm`)

`cd_ddmm()` tests whether a dataset's coordinate distribution is
consistent with a systematic "dd.mm → dd.dd" conversion error — a common
data-entry mistake where degree-minute values are incorrectly recorded as
decimal degrees (e.g., 45°30′ entered as 45.30° instead of 45.50°).

## Rasterized sampling (`cd_round`)

`cd_round()` tests whether a dataset's coordinates show an
unusually regular lattice pattern, which can indicate that points were
assigned to grid-cell centroids rather than actual sampling locations.

Both tests are parallelized by splitting datasets across workers (never
splitting a single dataset's records, which would break the
within-dataset statistics). All four calls (`cd_ddmm` diagnostic/flagged,
`cd_round` diagnostic/flagged) are wrapped in `tryCatch` because
`cd_ddmm()` can fail on datasets with very few records (its internal
`binom.test()` receives invalid inputs). When a chunk fails, its records
are conservatively kept (flagged as `TRUE`).

This step is **checkpointed**: if `dataset_level_flags.rds` already
exists, the saved flags are loaded.

```{r dataset-level-tests}
dataset_rds <- file.path(out_dir, "dataset_level_flags.rds")

if (file.exists(dataset_rds) && !FORCE_RERUN) {

  log_msg("Found existing dataset-level flags at ", dataset_rds, " -- loading from disk")
  saved <- readRDS(dataset_rds)
  ddmm_report  <- saved$ddmm_report
  round_report <- saved$round_report
  ddmm_flags   <- saved$ddmm_flags
  round_flags  <- saved$round_flags

} else {

  log_msg("Running cd_ddmm() and cd_round() across ", length(unique(clean$datasetKey)),
          " datasets (", N_WORKERS, " workers)...")

  dataset_ids <- unique(clean$datasetKey)
  dataset_chunks <- split(dataset_ids,
                          cut(seq_along(dataset_ids), N_WORKERS, labels = FALSE))

  chunk_results <- future_map(dataset_chunks, function(ds_chunk) {
    sub <- clean[clean$datasetKey %in% ds_chunk, ]

    ddmm_ds <- tryCatch({
      res <- cd_ddmm(sub, lon = "decimalLongitude", lat = "decimalLatitude",
                      ds = "datasetKey", diagnostic = TRUE, diff = 1, value = "dataset")
      tibble::rownames_to_column(res, var = "datasetKey")
    }, error = function(e) {
      message("  cd_ddmm() failed for a chunk (", length(ds_chunk), " datasets): ", conditionMessage(e))
      NULL
    })
    ddmm_fl <- tryCatch({
      cd_ddmm(sub, lon = "decimalLongitude", lat = "decimalLatitude",
              ds = "datasetKey", diff = 1, value = "flagged")
    }, error = function(e) rep(TRUE, nrow(sub)))

    round_ds <- tryCatch({
      cd_round(sub, lon = "decimalLongitude", lat = "decimalLatitude",
               ds = "datasetKey", T1 = 7, value = "dataset", graphs = FALSE)
    }, error = function(e) {
      message("  cd_round() failed for a chunk (", length(ds_chunk), " datasets): ", conditionMessage(e))
      NULL
    })
    round_fl <- tryCatch({
      cd_round(sub, lon = "decimalLongitude", lat = "decimalLatitude",
               ds = "datasetKey", T1 = 7, value = "flagged", graphs = FALSE)
    }, error = function(e) rep(TRUE, nrow(sub)))

    list(
      ddmm_report  = ddmm_ds,
      round_report = round_ds,
      flags        = tibble(row_id = sub$row_id, ddmm_flag = ddmm_fl, round_flag = round_fl)
    )
  }, .progress = FALSE)

  ddmm_report  <- bind_rows(map(chunk_results, "ddmm_report"))
  round_report <- bind_rows(map(chunk_results, "round_report"))
  flags_all    <- bind_rows(map(chunk_results, "flags")) %>% arrange(row_id)
  ddmm_flags   <- flags_all$ddmm_flag
  round_flags  <- flags_all$round_flag

  write_csv(ddmm_report, file.path(out_dir, "ddmm_conversion_flags_by_dataset.csv"))
  write_csv(round_report, file.path(out_dir, "rasterized_sampling_flags_by_dataset.csv"))

  saveRDS(list(ddmm_report = ddmm_report, round_report = round_report,
               ddmm_flags = ddmm_flags, round_flags = round_flags),
          dataset_rds)
}

clean_final <- clean[ddmm_flags & round_flags, ]

log_msg(nrow(clean) - nrow(clean_final),
        " records removed for belonging to a flagged dataset (ddmm/rasterized tests)")
log_msg(nrow(clean_final), " final cleaned records")
```

# Save outputs

Cleaned records are written to one CSV per species in
`gbif_clean_output/species_csvs/`. Column names `decimalLongitude` and
`decimalLatitude` are lowercased to `decimallongitude` and
`decimallatitude` for compatibility with downstream tools.

For species where GBIF's accepted name differs from the input name
(synonym cases identified [above](#backbone)), the output filename
includes both names:
`InputName_AcceptedName_Nn_YYYY-MM-DD.csv`.

```{r save-outputs}
clean_final <- clean_final %>%
  select(-row_id) %>%
  rename(decimallongitude = decimalLongitude, decimallatitude = decimalLatitude)

species_dir <- file.path(out_dir, "species_csvs")
dir.create(species_dir, showWarnings = FALSE)

for (sp in unique(clean_final$species)) {
  sp_data <- clean_final[clean_final$species == sp, ]

  input_match <- name_lookup$input_name[name_lookup$gbif_name == sp]
  if (length(input_match) == 1 && input_match != sp) {
    file_prefix <- paste0(gsub(" ", "_", input_match), "_", gsub(" ", "_", sp))
  } else {
    file_prefix <- gsub(" ", "_", sp)
  }

  sp_file <- file.path(species_dir, paste0(file_prefix, "_", nrow(sp_data), "n_",
                                           format(Sys.Date(), "%Y-%m-%d"), ".csv"))
  write_csv(sp_data, sp_file)
}

elapsed <- round(difftime(Sys.time(), script_start, units = "mins"), 1)
log_msg("=== Done in ", elapsed, " minutes. Cleaned data written to: ",
        species_dir, " (one CSV per species) ===")
```

# Reconciliation: missing species

Not every input species will necessarily have records in the final output.
This section identifies which species are missing and diagnoses why, by
checking whether they were present at each stage of the pipeline.

Species whose input name was a synonym (and whose records appear under
GBIF's accepted name) are excluded from the "missing" count — they do
have output, just under a different filename.

```{r reconciliation}
synonym_inputs <- name_lookup$input_name[name_lookup$gbif_name %in% unique(clean_final$species)]
missing_spp <- setdiff(species_list, c(unique(clean_final$species), synonym_inputs))
if (length(missing_spp) > 0) {
  log_msg(length(missing_spp), " species from the input list have no final output:")
  missing_info <- tibble(species = missing_spp) %>%
    mutate(
      in_gbif_download  = species %in% unique(dat_raw$species),
      after_cc_records   = species %in% unique(dat$species),
      after_cc_pipeline = species %in% unique(clean$species),
      after_precision   = species %in% unique(clean_final$species)
    )
  for (i in seq_len(nrow(missing_info))) {
    row <- missing_info[i, ]
    if (!row$in_gbif_download) {
      reason <- "no records returned by GBIF download (may be a synonym; check synonym_mapping.csv)"
    } else if (!row$after_cc_records) {
      reason <- "all records lacked coordinates"
    } else if (!row$after_cc_pipeline) {
      reason <- "all records removed by CoordinateCleaner tests (cc_val/cc_equ/.../cc_outl)"
    } else {
      reason <- "all records removed by precision or dataset-level filters"
    }
    log_msg("  - ", row$species, ": ", reason)
  }
  write_csv(missing_info, file.path(out_dir, "species_missing_from_output.csv"))
} else {
  log_msg("All ", length(species_list), " input species have records in the final output")
}
```
